Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-11-17 talk/pepm2011/paper.tex
comments from the review

The core idea of tracing JITs is to focus the optimization effort of the JIT
compiler on the commonly executed, \ie \emph{hot} paths of the core loops of the program and to just use an
interpreter for the less commonly executed parts. VMs that use a tracing JIT are
mostly mixed-mode execution environments, they contain both an interpreter and a
JIT compiler. By default the interpreter is used to execute the program, doing
some light-weight profiling at the same time. This profiling is used to identify
the hot loops of the program. If a hot loop is found in that way, the
interpreter enters a special \emph{tracing mode}. In this tracing mode, the
interpreter tries to record all operations that it is executing while running one
iteration of the hot loop. This history of executed operations of one loop is
called a \emph{trace}. Because the trace corresponds to one iteration of a loop,
it always ends with a jump to its own beginning. The trace also contains all
operations that are performed in functions that were called in the loop, thus a
tracing JIT automatically performs inlining.
This trace of operations subsequently forms the basis of the generated code. The trace is
first optimized, and then turned into machine code. Both optimization
and machine code generation are simple, because the traces are linear. This
linearity makes many optimizations a lot more tractable, and the inlining that
happens gives the optimizations automatically more context to work with.

@@

If one specific guard fails often enough, the tracing JIT will generate a new
trace that starts exactly at the position of the failing guard
\cite{andreas_gal_incremental_2006}. The existing assembler is patched to jump
to the new trace when the guard fails.  This approach guarantees that all the
hot paths in the program will eventually be traced and compiled into efficient
code.

@@

Using these classes to implement arithmetic shows the basic problem of a
dynamic language implementation. All the numbers are instances of either
\lstinline{BoxedInteger} or \lstinline{BoxedFloat}, therefore they consume space on the
heap. Performing many arithmetic operations produces lots of garbage quickly,
putthing pressure on the garbage collector. Using double dispatching to
implement the numeric tower needs two method calls per arithmetic operation,
which is costly due to the method dispatch.

@@

There exists a large number of works on escape analysis, which is a program
analysis that tries to find an upper bound for the lifetime of objects allocated
at specific program points
\cite{goldberg_higher_1990,park_escape_1992,choi_escape_1999,bruno_blanchet_escape_2003}.
This information can then be used to decide that certain objects can be
allocated on the stack, because their lifetime does not exceed that of the stack
frame it is allocated in. The difference to our work is that escape analysis is
split into an analysis and an optimization phase. The analysis can be a lot more
complex than our simple one-pass optimization. Also, stack-allocation reduces
garbage-collection pressure but does not optimize away the actual accesses to
the stack-allocated object. In our case, an object is not needed at all any
more.

@@

Partially known data structures are built directly into Prolog (via unbound
logic variables) and thus the treatment of partially static data structures
was part of partial evaluation of Prolog programs from the early stages
\cite{lloyd_partial_1991}. One effect of unfolding in Prolog is that terms that
are constructed and immediately matched again, completely disappear in the
residual program. This is similar to what our optimization does for an
imperative language.
In functional programming this idea was introduced as
 constructor specialisation by Mogensen \cite{mogensen_constructor_1993}.

>>>>

The core idea of tracing JITs is to focus the optimization effort of the JIT
compiler on the commonly executed, \ie \emph{hot} paths of the core loops of the program and to just use an
interpreter for the less commonly executed parts. VMs that use a tracing JIT are
mostly mixed-mode execution environments, they contain both an interpreter and a
JIT compiler. By default the interpreter is used to execute the program, doing
some light-weight profiling at the same time. This profiling is used to identify
the hot loops\reva{what is that? clarify} of the program. If a hot loop is found in that way, the
interpreter enters a special \emph{tracing mode}. In this tracing mode, the
interpreter tries to record all operations that it is executing while running one
iteration of the hot loop. This history of executed operations of one loop is
called a \emph{trace}. Because the trace corresponds to one iteration of a loop,
it always ends with a jump to its own beginning. The trace also contains all
operations that are performed in functions that were called in the loop, thus a
tracing JIT automatically performs inlining.
This trace of operations subsequently forms the basis of the generated code. The trace is
first optimized, and then turned into machine code. Both optimization
and machine code generation are simple, because the traces are linear. This
linearity makes many optimizations a lot more tractable, and the inlining that
happens gives the optimizations automatically more context to work with.

@@

If one specific guard fails often enough\reva{when is that?}, the tracing JIT will generate a new
trace that starts exactly at the position of the failing guard
\cite{andreas_gal_incremental_2006}. The existing assembler is patched to jump
to the new trace when the guard fails.  This approach guarantees that all the
hot paths in the program will eventually be traced and compiled into efficient
code.

@@

Using these classes to implement arithmetic shows the basic problem of a
dynamic language implementation. All the numbers are instances of either
\lstinline{BoxedInteger} or \lstinline{BoxedFloat}, therefore they consume space on the
heap. Performing many arithmetic operations produces lots of garbage quickly,
putting pressure on the garbage collector. Using double dispatching to
implement the numeric tower needs two method calls per arithmetic operation,
which is costly due to the method dispatch.

@@

There exists a large number of works on escape analysis, which is a program
analysis that tries to find an upper bound for the lifetime of objects allocated
at specific program points
\cite{goldberg_higher_1990,park_escape_1992,choi_escape_1999,bruno_blanchet_escape_2003}.
This information can then be used to decide that certain objects can be
allocated on the stack, because their lifetime does not exceed that of the stack
frame it is allocated in. The difference to our work is that escape analysis is
split into an analysis and an optimization phase. The analysis can be a lot more
complex than our simple one-pass optimization. Also, stack-allocation reduces
garbage-collection pressure but does not optimize away the actual accesses to
the stack-allocated object. In our case, an object is not needed at all any
more. \reva{Has this relation pointed out before in some paper? Is it a
novel contribution of your work?}

@@

Partially known data structures are built directly into Prolog (via unbound
logic variables) and thus the treatment of partially static data structures
was part of partial evaluation of Prolog programs from the early stages
\cite{lloyd_partial_1991}. One effect of unfolding in Prolog is that terms that
are constructed and immediately matched again, completely disappear in the
residual program. This is similar to what our optimization does for an
imperative language.
In functional programming this idea was introduced as
 constructor specialisation by Mogensen \cite{mogensen_constructor_1993}.
\reva{I'd refer here to
  deforestation (by Philip Wadler) and "short cut
  deforestation" by Andrew Gill, John Launchbury, and Simon L 
  Peyton Jones}
